State of the Art in Scientific Workflow Systems

### 2014-06-02 paper.tex
more refactor

\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%
\subsection{Architecture of a Generic SWE}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in an SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the
underlying \textit{distributed resources}. It resolves data dependencies and
optimizes the flow at runtime and monitors the workflow progress and collects
the intermediate and final results.

@@

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs in the previous decade, studying many existing systems for distributed
computing. This taxonomy presented is based, to a large extent on the
capabilities of the workflow engines. It does provide insights to many of the
SWE aspects considered in the present paper, especially the workflow languages
expressiveness and SWEs interactions with the distributed computing
infrastructures in the early days of the compute Grids. A more recent work by
Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely related to
the workflow language expressiveness: it compares the models of computations of
various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. A recent survey from
Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance
takes into account the latencies involved during the enactment mechanisms. A
small survey and analysis of file-access characteristics for data-intensive
workflows has been done by Shibata~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in~\cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR
DAGMan~\cite{dagman} to abstractions for parallel computations such as Petri
nets~\cite{alt-hoheisel:2005}, data-driven languages such as
Scufl~\cite{turi-missier-etal:2007} and scripting languages such as Swift
script~ \cite{swift}. Glatard's PhD thesis~\cite{glatard:2007} presents a
taxonomy of SWE languages classifying them into functional, graph,
service-oriented and executable representations. A detailed classification and
taxonomy of parallelism patterns in Grid Workflows has been presented by
Pautasso and Alonso in~\cite{pautasso1}. This classification is more generic
while the one presented in Glatard's thesis has emphasis on service based
workflows.  Each of these approaches can be defended through some aspects well
covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of application logic for non-expert users and scripting is
extensively used by programmers for prototyping, etc.

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field

% Say that different workflow systems have different design goals. It is hard
% to chose one from the existing or build your own. It is almost equally
% complex to build own just as it is complex to create own programming
% platform. We hope to provide by the way of this paper a guide for new adopters on how to chose a workflow language appropriate for their needs. We ask and answer the following questions of the workflow systems that we study in this paper:
%-- What were the original design goals?
%-- What are the main application areas that the system is currently being used?
%-- What are the characteristics and capabilities of the language? expressivity features?
%-- What are the characteristics and capabilities of the run-time? parallelism? concurrency? task and data grouping?
%-- What kinds of computing and computational infrastructure does the workflow system cater to? Grids, Clouds, Clusters, Supercomputers, fast IB networks? Accelerators?
%-- How have they evolved? What are the possible future directions?

In the present work we survey and study the various contemporary SWEs in order
to find the capabilities and key features in terms of the characteristics
described in section~\ref{sec:intro}. We believe this study will serve to
address the following objectives:

\begin{enumerate}
\item \textbf{Educate.} With more science being done with the help of
computational methods, SWEs can add significant value to the process.  They can
accelerate the process, enable community driven science, drive the development
at a manageable level of abstraction for users.  However, scientific users are
often not experienced programmers or computation savvy. This work will serve as
a useful starting guide to educate science users and students of the available
tools and techniques with features suitable for their methods.

\item \textbf{Catalog.} We study a number of important features as offered by
the contemporary SWEs. This will help catalog and collect the SWEs by their
offerings and identify mechanisms and needs for new and existing features
in one place. This will also help identify current trends and approaches
to computation in scientific communities.

\item \textbf{Application.} We identify the applications and application
classes well adapted to the workflow systems and are benefitting by them.
This will help readers relate to the application they are associated to.
We hope that readers can make a choice of which SWE to use if there is an
adaptation similar in scope to the ones describe in this paper.

\item \textbf{Evolution.} We comment on the evolution of the existing workflow
systems and the possible current and future development directions which
will help readers be aware of the general development direction of the
SWEs.
\end{enumerate}

The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work.

Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work. We rely on the following sources for our information: 
\begin{enumerate}
    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
    \item \textbf{Technical Reports.} Technical reports and theses.
    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
\end{enumerate}

Some features, such as extendability of an SWE strongly depend on the extension
points designed in the code; consequently, we consider it present in only those
SWEs that have an explicit documented description or demonstration of the given
feature.

@@

\input{text/taverna}
\input{text/moteur}
\input{text/makeflow}
\input{text/triana}
\input{text/kepler}
\input{text/peg}
\input{text/wings}
\input{text/swift}
\input{text/askalon}
\input{text/pgrade}
\input{text/galaxy}
% Put this somewhere else, either in the analysis or may be in the conclusions
% and summary.
%The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
%is similar in spirit to the
%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A Swift script 
%contains statically typed datasets that can be declared
%in the script. Unlike Swift script, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
%represented using XML-representation. This means that effectively, a
%workflow is a DAG. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. Conditional and Loop execution is supported
%using the if and foreach statements in both languages.
%
%

@@

The software architecture of an enactor plays a key role in determining the
scalability for the workflow applications exhibiting a large number of tasks.
An architecture that takes into account the properties of the underlying
infrastructure could scale well in the face of growing number of workflow
tasks.

@@

In this section we analyze the scalability measures and techniques employed by
contemporary SWEs. Illustrated below are some of the techniques used to achieve
scalability in various SWEs.

>>>>

\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%

\subsection{Architecture}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in an SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the
underlying \textit{distributed resources}. It resolves data dependencies and
optimizes the flow at runtime and monitors the workflow progress and collects
the intermediate and final results.

@@

\input{text/related}
\input{text/objectives}

@@

\input{text/taverna}
\input{text/moteur}
\input{text/makeflow}
\input{text/triana}
\input{text/kepler}
\input{text/peg}
\input{text/wings}
\input{text/swift}
\input{text/askalon}
\input{text/pgrade}
\input{text/galaxy}
% Put this in analysis section. 
%The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
%is similar in spirit to the SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A Swift script 
%contains statically typed datasets that can be declared
%in the script. Unlike Swift script, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
%represented using XML-representation. This means that effectively, a
%workflow is a DAG. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. Conditional and Loop execution is supported
%using the if and foreach statements in both languages.
%
%

@@

The software architecture of the system plays a key role in determining the
scalability for the workflow applications exhibiting a large number of tasks.
An architecture that takes into account the properties of the underlying
infrastructure could scale well in the face of growing number of workflow
tasks.

@@

In this section we analyze the scalability and runtime optimization measures
and techniques employed by SWEs. Illustrated below are some of the techniques
used to achieve scalability in various SWEs.
